Effect of quality-adjusted production function inputs

### 2015-05-22 Articles/Presentation_ESEE_2015/ESEE_2015.Rnw
All Solow residual derivations in appendix

%..............................
\paragraph{Estimating the Solow residual} 
\label{sec:estimating_the_solow_residual}
%..............................

Assuming that parameters $\theta$, $\alpha$, and $\beta$
in Equation~\ref{eq:CDe} are known from parameter estimation,
we can estimate the value of $\lambda$ as follows.
First, assume constant returns to scale such that 
$\alpha + \beta + \gamma = 1$, 
and calculate $\gamma = 1 - \alpha - \beta$.
Then, take the natural logarithm ($\ln$) of Equation~\ref{eq:CDe} to obtain
%
\begin{equation} \label{eq:lnCD}
  \ln y = \ln \theta + \ln A + \alpha \ln k + \beta \ln l + (1 - \alpha - \beta) \ln e \;.
\end{equation}
%
By taking the derivative of Equation~\ref{eq:lnCD} with respect to time ($t$) 
and noting that model parameters $\theta$, $\alpha$, and $\beta$ are 
constant with respect to time, we obtain
%
\begin{equation} \label{eq:ddtlnCD}
  \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      = \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      + \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} \;.
\end{equation}
%
Solving for the total factor productivity term gives
%
\begin{equation} \label{eq:solow_resid}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
        \right] \;.
\end{equation}
%
Recognizing that $A \equiv \mathrm{e}^{\lambda (t-t_0)}$ gives
%
\begin{equation}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
    = \frac{1}{\mathrm{e}^{\lambda (t-t_0)}} \; \lambda \mathrm{e}^{\lambda (t-t_0)}
    = \lambda \; ,
\end{equation}
%
which can be substituted into Equation~\ref{eq:solow_resid} to find
%
\begin{equation} \label{eq:lambda}
  \lambda 
    = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
    - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
      \right] \;.
\end{equation}
%
Equation~\ref{eq:lambda} applies for any instant in time.

%..............................
\paragraph{Fraction of growth explained by the Solow residual} 
\label{sec:fraction_of_growth_solow}
%..............................

The fraction~($f$) of GDP growth explained by the Solow residual
can be given as
%
\begin{equation} \label{eq:frac_growth_explained_by_solow}
  f \equiv \frac{\lambda}{\frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t}} 
            = 1 - \frac{\alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
                      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
                      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t}}
                  {\frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t}} \; .
\end{equation}
%
Using forward differences to estimate $f$ yields
%
\begin{equation} \label{eq:f_finite_diff_1}
  f_{i,j} = \frac{\lambda_{i,j}}{\frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i}}
       = 1 - \frac{\alpha \frac{1}{k_i}\frac{k_j - k_i}{t_j - t_i} 
            + \beta \frac{1}{l_i}\frac{l_j - l_i}{t_j - t_i} 
            + (1 - \alpha - \beta) \frac{1}{e_i}\frac{e_j - e_i}{t_j - t_i} }
            {\frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i}} \; .
\end{equation}
%
Cancelling $t_j - t_i$ terms and simplifying gives
%
\begin{equation} \label{eq:f_finite_diff_2}
  f_{i,j} = 1 - \frac{\alpha \left( \frac{k_j}{k_i} - 1 \right) 
                      + \beta \left( \frac{l_j}{l_i} - 1 \right)
                      + (1 - \alpha - \beta) \left( \frac{e_j}{e_i} - 1 \right)}
            {\left( \frac{y_j}{y_i} - 1 \right)} \; ,
\end{equation}
%
which is an estimate of the instantaneous value of $f$ 
given by Equation~\ref{eq:frac_growth_explained_by_solow}.

@@

<<Models>>=
fileName <- "ESEE2015Models.Rdata"
if (!rerunModels){
  # load from the file on disk
  load(fileName)
} else {
  # re-run the models and save to disk.
  # In RStudio, be sure to set the working directory to the source file directory.
  flavors <- c("unadjusted", "adjusted")
  nests <- NULL
  nests$kle <- c(1,2,3); nests$lek <- c(2,3,1); nests$ekl <- c(3,1,2)

@@

flavors <- c("unadjusted", "adjusted")
# model.types <- c("CD", "CES")
model.types <- c("CD")
nests <- list(kle=c(1,2,3), lek=c(2,3,1), ekl=c(3,1,2))

dframes <- list()
for (Coun in countries){
  for (flav in flavors){
    forms <- list(noE = keep_summands(formulas[[Coun]][[flav]], -3), withE = formulas[[Coun]][[flav]])
    for (form in forms){
      denom <- summands(form, n=length(summands(form))) # time variable
      nums <- all.vars(form) [1:(length(all.vars(form))-1)] # y, k, l, and e (if present)
      trimmedData <- historicalData[[Coun]][c(denom, nums)] # columns are: iYear, y, k, l, and e (if present)
      df <- historicalData[[Coun]][ , c(nums, "Year")]
      # Calculate names for the columns of finite difference derivatives
      derivNames <- lapply(1:length(nums), function(i){paste0("d", nums[[i]], "/", "dt")}) %>% unlist
      derivCols <- lapply(1:length(nums), function(i){
        # Calculate finite difference derivative columns
        ediff(trimmedData[[i+1]], pad=pad, lag=lag) / ediff(trimmedData[[1]], pad=pad, lag=lag)
      }) %>% do.call(cbind, .) %>% as.data.frame
      names(derivCols) <- derivNames
      df <- cbind(derivCols, df)
      e <- if (length(summands(form))==4) "withE" else "noE"
      for (m in model.types){
        origCoeffs <- filter(AllCoeffs, model==m & energy==e & Country==Coun & flavor==flav)
        df <- cbind(df, origCoeffs)
        if (m == "CD"){
          lambda <- lapply(1:length(derivNames), function(i){
            # derivative term divided by its numerator.
            # derivative term is in the ith column of df
            # numerator is in i + length(derivNames) th column of df
            out <- df[[i]] / df[[i+length(derivNames)]]
            if (i > 1){
              # Need to make it negative and multiply by the corresponding alpha term.
              # Typical setup is that columns 2 is capital, and we need alpha_1 for capital.
              # So, subtract 1 from i.
              out <- - out * df[[paste0("alpha_", i-1)]]
            }
            return(out)
          }) %>% as.data.frame %>% rowSums
          df$lambda_fd <- lambda
          avg.lambda_fd <- mean(df$lambda_fd, na.rm = TRUE)
          # Fraction of growth explained by Solow Residual
          df$solowFrac <- df$lambda_fd / (df[[1]] / df[[length(derivNames) + 1]])
          avg.solowFrac <- mean(df$solowFrac, na.rm = TRUE)
        } else if (m == "CES"){

@@

Figure~\ref{fig:DynamicSR_CD_Graph} shows the evolution 
of the Solow residual~($\lambda$) over time
as calculated by Equation~\ref{eq:lambdaij} 
with the timespan for finite differences ($t_j - t_i$) set to 
\Sexpr{lag}~\Sexpr{if (lag == 1) "year" else "years"}.
It appears that quality-adjusting the factors of production
decreases the Solow residual~($\lambda$).

@@

\section{Appendix}

Include data tables here?

>>>>

@@

<<Models>>=
flavors <- c("unadjusted", "adjusted")
nests <- list(kle=c(1,2,3), lek=c(2,3,1), ekl=c(3,1,2))
model.types <- c("CD", "CES")

fileName <- "ESEE2015Models.Rdata"
if (!rerunModels){
  # load from the file on disk
  load(fileName)
} else {
  # re-run the models and save to disk.
  # In RStudio, be sure to set the working directory to the source file directory.

@@

dframes <- list()
for (Coun in countries){
  for (flav in flavors){
    forms <- list(noE = keep_summands(formulas[[Coun]][[flav]], -3), withE = formulas[[Coun]][[flav]])
    for (form in forms){
      denom <- summands(form, n=length(summands(form))) # time variable
      nums <- all.vars(form) [1:(length(all.vars(form))-1)] # y, k, l, and e (if present)
      df <- historicalData[[Coun]][c(denom, nums)] # columns are: iYear, y, k, l, and e (if present)
      # Calculate names for the columns of finite difference derivatives
      derivNames <- lapply(1:length(nums), function(i){paste0("d", nums[[i]], "/", "dt")}) %>% unlist
      derivCols <- lapply(1:length(nums), function(i){
        # Calculate finite difference derivative columns
        ediff(df[[i+1]], pad=pad, lag=lag) / ediff(df[[1]], pad=pad, lag=lag)
      }) %>% do.call(cbind, .) %>% as.data.frame
      names(derivCols) <- derivNames
      df <- cbind(derivCols, historicalData[[Coun]][ , c(nums, "Year")])
      e <- if (length(summands(form))==4) "withE" else "noE"
      for (m in model.types){
        origCoeffs <- filter(AllCoeffs, model==m & energy==e & Country==Coun & flavor==flav)
        df <- cbind(df, origCoeffs)
        if (m == "CD"){
          lambda <- lapply(1:length(derivNames), function(i){
            # derivative term divided by its numerator.
            # derivative term is in the ith column of df
            # numerator is in i + length(derivNames) th column of df
            out <- df[[i]] / df[[i+length(derivNames)]]
            if (i > 1){
              # Need to make it negative and multiply by the corresponding alpha term.
              # Typical setup is that columns 2 is capital, and we need alpha_1 for capital.
              # So, subtract 1 from i.
              out <- - out * df[[paste0("alpha_", i-1)]]
            }
            return(out)
          }) %>% as.data.frame %>% rowSums
          df$lambda_fd <- lambda
          avg.lambda_fd <- mean(df$lambda_fd, na.rm = TRUE)
          # Fraction of growth explained by Solow Residual
          df$solowFrac <- df$lambda_fd / (df[[1]] / df[[length(derivNames) + 1]])
          avg.solowFrac <- mean(df$solowFrac, na.rm = TRUE)
        } else if (m == "CES"){
          if (e == "noE"){
            # kl nesting
#             df$lambda_fd <- df[[1]] / 
          } else if (e == "withE"){

          } else {
            stop (paste("Unknown e:", e))
          }

@@

Figure~\ref{fig:DynamicSR_CD_Graph} shows the evolution 
of the Solow residual~($\lambda$) over time
as calculated by Equation~\ref{eq:lambdaij} 
with the timespan for finite differences ($t_j - t_i$) set to 
\Sexpr{lag}~\Sexpr{if (lag == 1) "year" else "years"}.
It appears that quality-adjusting the factors of production
\emph{and} adding augmenting the production function with energy
decreases the Solow residual~($\lambda$),
i.e., brings the Solow residual closer to zero.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Derivation of dynamic Solow residual for the CD equation}
\label{sec:dynamic_sr_CD}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Assuming that parameters $\theta$, $\alpha$, and $\beta$
in Equation~\ref{eq:CDe} are known from parameter estimation,
we can estimate the value of $\lambda$ as follows.
First, assume constant returns to scale such that 
$\alpha + \beta + \gamma = 1$, 
and calculate $\gamma = 1 - \alpha - \beta$.
Then, take the natural logarithm ($\ln$) of Equation~\ref{eq:CDe} to obtain
%
\begin{equation} \label{eq:lnCD}
  \ln y = \ln \theta + \ln A + \alpha \ln k + \beta \ln l + (1 - \alpha - \beta) \ln e \;.
\end{equation}
%
By taking the derivative of Equation~\ref{eq:lnCD} with respect to time ($t$) 
and noting that model parameters $\theta$, $\alpha$, and $\beta$ are 
constant with respect to time, we obtain
%
\begin{equation} \label{eq:ddtlnCD}
  \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      = \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      + \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} \;.
\end{equation}
%
Solving for the total factor productivity term gives
%
\begin{equation} \label{eq:solow_resid}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
      = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
      - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
        \right] \;.
\end{equation}
%
Recognizing that $A \equiv \mathrm{e}^{\lambda (t-t_0)}$ gives
%
\begin{equation} \label{eq:lambda_calc}
  \frac{1}{A}\frac{\mathrm{d}A}{\mathrm{d}t} 
    = \frac{1}{\mathrm{e}^{\lambda (t-t_0)}} \; \lambda \mathrm{e}^{\lambda (t-t_0)}
    = \lambda \; ,
\end{equation}
%
which can be substituted into Equation~\ref{eq:solow_resid} to find
%
\begin{equation} \label{eq:lambda}
  \lambda 
    = \frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t} 
    - \left[
              \alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
              + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
              + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t} 
      \right] \;.
\end{equation}
%
Equation~\ref{eq:lambda} applies for any instant in time.

The fraction~($f$) of GDP growth explained by the Solow residual
for the CD model
can be given as
%
\begin{equation} \label{eq:frac_growth_explained_by_solow}
  f \equiv \frac{\lambda}{\frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t}} 
            = 1 - \frac{\alpha \frac{1}{k}\frac{\mathrm{d}k}{\mathrm{d}t} 
                      + \beta \frac{1}{l}\frac{\mathrm{d}l}{\mathrm{d}t} 
                      + (1 - \alpha - \beta) \frac{1}{e}\frac{\mathrm{d}e}{\mathrm{d}t}}
                  {\frac{1}{y}\frac{\mathrm{d}y}{\mathrm{d}t}} \; .
\end{equation}
%
Using forward differences to estimate $f$ yields
%
\begin{equation} \label{eq:f_finite_diff_1}
  f_{i,j} = \frac{\lambda_{i,j}}{\frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i}}
       = 1 - \frac{\alpha \frac{1}{k_i}\frac{k_j - k_i}{t_j - t_i} 
            + \beta \frac{1}{l_i}\frac{l_j - l_i}{t_j - t_i} 
            + (1 - \alpha - \beta) \frac{1}{e_i}\frac{e_j - e_i}{t_j - t_i} }
            {\frac{1}{y_i}\frac{y_j - y_i}{t_j - t_i}} \; .
\end{equation}
%
Cancelling $t_j - t_i$ terms and simplifying gives
%
\begin{equation} \label{eq:f_finite_diff_2}
  f_{i,j} = 1 - \frac{\alpha \left( \frac{k_j}{k_i} - 1 \right) 
                      + \beta \left( \frac{l_j}{l_i} - 1 \right)
                      + (1 - \alpha - \beta) \left( \frac{e_j}{e_i} - 1 \right)}
            {\left( \frac{y_j}{y_i} - 1 \right)} \; ,
\end{equation}
%
which is an estimate of the instantaneous value of $f$ 
given by Equation~\ref{eq:frac_growth_explained_by_solow}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Derivation of dynamic Solow residual for the CES equation}
\label{sec:dynamic_sr_CES}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In this section, we derive the dynamic Solow residual ($\lambda$) 
for the CES equation with three generic factors of production, namely
$x_1$, $x_2$, and $x_3$.

We can define
%
\begin{equation} \label{eq:a_def}
  a \equiv \delta b^{\rho/\rho_1} + (1-\delta) x_3^{-\rho}
\end{equation}
%
and
%
\begin{equation} \label{eq:b_def}
  b \equiv \delta_1 x_1^{-\rho_1} + (1 - \delta_1) x_2^{-\rho_1} \; ,
\end{equation}
%
such that Equation~\ref{eq:CESgeneric} can be restated as
%
\begin{equation} \label{eq:simple_y} 
  y = \gamma A a^{-1/\rho} \; .
\end{equation}
%
Taking the natural logarithm of Equation~\ref{eq:simple_y}
and realizing that $\gamma$ is not a function of time, we find
%
\begin{equation}
  \frac{1}{y} \frac{\mathrm{d} y}{\mathrm{d} t} = \frac{1}{A} \frac{\mathrm{d} A}{\mathrm{d} t} 
                      + \left( - \frac{1}{\rho} \right) \frac{1}{a} \frac{\mathrm{d} a}{\mathrm{d} t} \; .
\end{equation}
%
By Equation~\ref{eq:lambda_calc}, and after rearranging, 
the dynamic Solow residual for the CES equation can be stated as
%
\begin{equation} \label{eq:lambda_for_CES}
  \lambda = \frac{1}{y} \frac{\mathrm{d} y}{\mathrm{d} t} + \frac{1}{a} \frac{1}{\rho} \frac{\mathrm{d} a}{\mathrm{d} t} \; .
\end{equation}
%
To find $\frac{\mathrm{d} a}{\mathrm{d} t}$ and $\frac{\mathrm{d} b}{\mathrm{d} t}$, 
we take the time derivatives 
of Equations~\ref{eq:a_def} and~\ref{eq:b_def}
and rearrange slightly to obtain
%
\begin{equation} \label{eq:da_dt}
  \frac{1}{\rho} \frac{\mathrm{d} a}{\mathrm{d} t} = \delta b^{(\rho/\rho_1 - 1)} \frac{1}{\rho_1} \frac{\mathrm{d} b}{\mathrm{d} t}
                                      - (1-\delta) x_3^{-(\rho + 1)} \frac{\mathrm{d} x_3}{\mathrm{d} t}
\end{equation}
%
and
%
\begin{equation} \label{eq:db_dt}
  \frac{1}{\rho_1} \frac{\mathrm{d} b}{\mathrm{d} t} = - \delta_1 x_1^{-(\rho_1 + 1)} \frac{\mathrm{d} x_1}{\mathrm{d} t}
                                      - (1 - \delta_1) x_2^{-(\rho_1 + 1)} \frac{\mathrm{d} x_2}{\mathrm{d} t} \; .
\end{equation}
%
Equations~\ref{eq:lambda_for_CES}--\ref{eq:db_dt} can be used to 
calculate a dynamic time series for $\lambda$ given 
fitted parameters from the CES model
($\rho, \rho_1, \delta, \, \mathrm{and} \; \delta_1$)
and time series for economic output~($y$) and
factors of production 
($x_1, x_2, \, \mathrm{and} \; x_3$).
The time derivatives $\left( 
\frac{\mathrm{d} y}{\mathrm{d} t}, 
\frac{\mathrm{d} x_1}{\mathrm{d} t},
\frac{\mathrm{d} x_2}{\mathrm{d} t}, \, \mathrm{and} \;
\frac{\mathrm{d} x_3}{\mathrm{d} t} \right)$
can be approximated from historical data
by forward finite differences in a manner similar to Equation~\ref{eq:lambdaij}.
